Predicted cardiovascular risk for US adults with diabetes, chronic kidney disease, and \u2265 65 years of age

### 2020-10-29 doc/ACCAHA_BP_groups.Rmd
second draft feedback incorporated

Byron C. Jaeger, PhD^1^, Swati Sakhuja, MS?^2^, Shakia T. Hardy, PhD^2^, Oluwasegun P. Akinyelure, MD, MPH^2^, Joshua D. Bundy, PhD, MPH^3^, Paul Muntner, PhD^2^, and Paul K. Whelton, MD, MSc^3^

@@

__Funding:__ Information on grants, contracts, and other forms of financial support. _Please add your data if needed_

@@

Dr. Whelton was supported by a Centers for Research Excellence grant from the National Institute of General Medical Sciences (P20GM109036)

__Disclosures:__ State each author's disclosures (or lack thereof). This must include the full disclosure of any relationship with industry. _Please add your statement below_

BCJ has nothing to disclose. 
OPA has nothing to disclose. 
JDB has nothing to disclose.
STH has nothing to disclose.
PKW has nothing to disclose.

__Address for correspondence:__
Byron C. Jaeger
University of Alabama at Birmingham
327M Ryals Public Health Building
1665 University Blvd
Birmingham, Alabama 35294-0022

@@

_Note: Maximum word count is 250_

@@

__Objectives:__ Determine if the majority of US adults in subgroups defined by diabetes, chronic kidney disease (CKD), and age ≥65 years have high ASCVD risk (i.e., 10-year predicted ASCVD risk ≥10% or clinical CVD), and estimate the age-adjusted probability of high ASCVD risk.

__Methods:__ Adults aged 40-79 years from the National Health and Nutrition Examination Survey `r min(exams)`-`r max(exams)+1` were included (n=`r table_value(nrow(current_analysis$data))`). We predicted 10-year ASCVD risk using the Pooled Cohort risk equations and clinical CVD was self-reported. Analyses were conducted overall and among those with stage 1 hypertension, defined by systolic BP of 130-139 mmHg or diastolic BP of 80-89 mm Hg.

@@

_Note: No more than 100 words, stressing clinical implications_

@@

Distribution of US adults by blood pressure and antihypertensive medication use categories. In text, the proportion of adults with stage 1 hypertension at high atherosclerotic cardiovascular disease risk in subgroups defined by diabetes, chronic kidney disease and age ≥ 65 years.
`r fp_par(line_spacing = 1)`

@@

```{r fig.width=7, fig.height = 4, dpi = 300}

fig_central_illustration

@@

High atherosclerotic cardiovascular disease risk was defined by a 10-year predicted risk for atherosclerotic cardiovascular disease ≥ 10% or clinical cardiovascular disease.
`r fp_par(line_spacing = 1)`

@@

__Translational Outlook:__ Predicted 10-year ASCVD risk should be computed for all adults with stage 1 hypertension and without clinical CVD as many are not at high risk for ASCVD.

@@

In November 2017, the `r ACC_AHA()` published a guideline for the prevention, detection, evaluation, and management of high `r BP()` in adults [@whelton20182017]. This guideline recommends using both `r BP()` levels and predicted risk for `r CVD()` to guide the initiation of antihypertensive medication in addition to nonpharmacological therapy. All adults with an average `r SBP()` ≥ 140 mm Hg or `r DBP()` ≥ 90 mm Hg, stage 2 hypertension in the guideline, are recommended to initiate antihypertensive medication. Adults with `r SBP()` between 130 and 139 mm Hg and/or `r DBP()` between 80 and 89 mm Hg, stage 1 hypertension in the guideline, with high `r CVD()` risk are recommended to initiate antihypertensive medication. High `r CVD()` risk was defined by the presence of clinical `r CVD()` or a 10-year predicted risk for `r ASCVD()` ≥ 10% [@goff20142013]. Additionally, adults with with stage 1 hypertension and diabetes mellitus, `r CKD()`, or ≥ 65 years of age are considered to have high CVD risk by the guideline and are recommended to initiate antihypertensive medication. This recommendation, written in Sections 9.3, 9.6, and 10.3 of the guideline, was based on the assumption that the vast majority of adults with diabetes, `r CKD()`, or ≥ 65 years of age are likely to have a 10-year predicted `r ASCVD()` risk ≥ 10% [@whelton20182017].

@@

The current analysis was restricted to adults aged 40 to 79 years of age who completed the `r NHANES()` interview and examination (n = `r n_age_40_to_79`). Participants < 40 or > 79 years of age were not included because use of the Pooled Cohort risk equations is not recommended to be used in these age ranges [@goff20142013]. Participants who did not have three `r SBP()` and `r DBP()` measurements (n = `r n_bp_too_low`) and those who were missing information on age, race, sex, total and high-density lipoprotein cholesterol, smoking status, diabetes, or `r CKD()` status (n = `r n_missing_pcr`) were excluded. After these exclusions, a total of `r n_final` participants were included in the analysis (`r refer_to('fig_include_exclude')`).

@@

Data were collected during an in-home interview and a study visit completed at a mobile examination center. Standardized questionnaires were used to assess participants’ age, sex, race/ethnicity, smoking habits, medical history, use of antihypertensive medication, oral glucose lowering medication, and insulin. The medical history assessment included questions about whether the participant had been told by a doctor or other health professional that they had a heart attack, coronary heart disease stroke, or heart failure. Clinical `r CVD()` was defined as answering yes to at least one of these questions.

@@

Blood and urine samples were collected during the medical examination. Serum creatinine, serum glucose and `r HB_A1C()` were measured using standard methods. Diabetes was defined by fasting plasma glucose ≥ `r gluc_cutpoint_fasted` mg/dL (≥ `r gluc_cutpoint_fed` mg/dL for those who were not fasting), `r HB_A1C()` ≥ `r hba1c_cutpoint`%, or self-reported use of insulin or oral glucose lowering medication. `r capitalize(E_GFR())` was calculated using the `r CKD()` Epidemiology Collaboration equation [@levey2009new; @cite_nephro]. Urinary albumin and creatinine levels were measured and used to calculate the `r ACR()`. `r CKD()` was defined by an `r E_GFR()` < `r egfr_cutpoint` ml/min/1.73m^2^ or an `r ACR()` ≥ `r acr_cutpoint` mg/g. Ten-year predicted `r ASCVD()` risk was calculated using the Pooled Cohort risk equations for participants without clinical `r CVD()` [@goff20142013]. High `r ASCVD()` risk was defined as clinical `r CVD()` or a 10-year predicted `r ASCVD()` risk ≥ 10%.

@@

Among US adults aged 40 to 79 years, the estimated percentage (95% `r CI()`) with high `r ASCVD()` risk was `r inline$prevHighRiskOverall` (`r refer_to('tab_risk_overall')`; bottom panel). Among sub-groups with diabetes, `r CKD()`, age ≥ 65 years, or any of these conditions, `r inline$prevHighRiskDiabetes`, `r inline$prevHighRiskCkd`, `r inline$prevHighRiskAge65`, and `r inline$prevHighRiskAny` had high `r ASCVD()` risk, respectively . Among those with stage 1 hypertension, the estimated percentage (95% `r CI()`) with high `r ASCVD()` risk was `r inline$prevHighRiskS1hOverall` in the overall population and `r inline$prevHighRiskS1hDiabetes`, `r inline$prevHighRiskS1hCkd`, `r inline$prevHighRiskS1hAge65`, and `r inline$prevHighRiskS1hAny` for those with diabetes, `r CKD()`, age ≥ 65 years, or any of these conditions, respectively.

@@

In the current study, a majority of US adults aged 40 to 79 years with diabetes, `r CKD()`, age ≥ 65 years, and any of these conditions had high `r ASCVD()` risk, defined by a 10-year predicted `r ASCVD()` risk ≥ 10% or clinical CVD. Also, the majority of US adults with stage 1 hypertension who were ≥ 65 years of age had high `r ASCVD()` risk. However, a substantial proportion of US adults with stage 1 hypertension and diabetes or `r CKD()` did not have high `r ASCVD()` risk. Among US adults with `r CKD()` and without high `r ASCVD()` risk, approximately one-third had a 10-year predicted `r ASCVD()` risk < 2.5%. In contrast, over 80% of US adults aged ≥ 65 years without high `r ASCVD()` risk had a 10-year predicted `r ASCVD()` risk between 5% and 10%. The probability of having high `r ASCVD()` risk was age-dependent with over 50% of US adults with diabetes and `r CKD()` having high `r ASCVD()` risk at ages above `r inline$ageHighRiskOvrlDiabetes` and `r inline$ageHighRiskOvrlCkd` years, respectively.

@@

Previous randomized trials and meta-analyses have investigated whether lower `r BP()` goals reduce incident `r CVD()` risk in patients with diabetes, `r CKD()`, or older age. A randomized trial of patients with diabetes and baseline `r SBP()` < 140 mm hg and `r DBP()` < 80 mm Hg found lower risk of stroke among participants with more intense `r BP()` treatment goals [@schrier2002effects]. The Systolic Blood Pressure Intervention Trial compared an `r SBP()` treatment target of < 120 mm Hg to < 140 mm Hg among older adults (mean age of 68 years) without diabetes. Among participants with `r CKD()`, randomization to the lower `r SBP()` target reduced rates of major `r CVD()` events and all-cause death without evidence of effect modifications by `r CKD()` or deleterious effect on `r E_GFR()` or end stage renal disease [@cheung2017effects]. In the overall population, those randomized to the lower `r SBP()` target experienced lower rates of incident fatal and nonfatal major `r CVD()` events as well as all-cause mortality [@sprint2015randomized]. A systematic review and meta-analysis of 123 studies with 613,815 participants found strong support for lowering `r SBP()` to < 130 mm Hg and providing antihypertensive medication to adults with diabetes, CKD, and various other comorbidities [@ettehad2016blood]. Benefits associated with a `r SBP()` treatment goal of < 130 mm Hg have also been identified in several other meta-analyses [@bundy2017systolic; @reboussin2018systematic]. Collectively, evidence from previous studies suggests that adults with hypertension and diabetes, `r CKD()` or ≥ 65 years of age may experience lower rates of `r CVD()` events by initiating antihypertensive medication with an `r SBP()` treatment goal of < 130 mm Hg versus < 140 mm Hg.

@@

It has been suggested that the vast majority of adults with diabetes, CKD, or ≥ 65 years of age have a 10-year predicted `r ASCVD()` risk ≥ 10% [@whelton20182017]. Previous research has shown that diabetes, `r CKD()`, and older age are each associated with an increased risk for `r ASCVD()` events [@coresh2004evidence; @chronic2010association; @grundy1999diabetes; @lakatta2002age]. Although the current study suggests that a high proportion of US adults with stage 1 hypertension and diabetes or `r CKD()` do not have a 10-year predicted `r ASCVD()` risk ≥ 10%, diabetes and `r CKD()` are associated with a high lifetime `r CVD()` risk [@lloyd2006prediction; @hippisley2010derivation]. Age-specific estimates in the current study suggest that US adults with diabetes or `r CKD()` develop high `r ASCVD()` risk at a younger age than their counterparts without these conditions. Prior studies have also found that cumulative exposure to high `r BP()` is associated with increased `r CVD()` risk [@allen2014blood]. Therefore, for younger adults with diabetes or `r CKD()`, early initiation of antihypertensive medication may be an important step towards lowering lifetime `r CVD()` risk. Estimating lifetime CVD risk may be useful when discussing the initiation of antihypertensive medication among young and middle-aged adults with diabetes or CKD.

>>>>

Byron C. Jaeger, PhD^1^, Swati Sakhuja, MPH^2^, Shakia T. Hardy, PhD, MPH^2^, Oluwasegun P. Akinyelure, MD, MPH^2^, Joshua D. Bundy, PhD, MPH^3^, Paul Muntner, PhD^2^, and Paul K. Whelton, MD, MSc^3^

@@

__Funding:__

@@

Dr. Bundy receives support through K12HD043451 from the Eunice Kennedy Shriver National Institute of Child Health and Human Development.

Dr. Whelton receives support through P20GM109036 from the Centers for Research Excellence grant from the National Institute of General Medical Sciences.

__Disclosures:__

BCJ has nothing to disclose. 
SS has nothing to disclose.
STH has nothing to disclose.
OPA has nothing to disclose. 
JDB has nothing to disclose.
PM received grant funding and consulting fees from Amgen Inc., unrelated to the current manuscript.
PKW has nothing to disclose.

__Address for correspondence:__

Byron C. Jaeger
University of Alabama at Birmingham
327M Ryals Public Health Building
1665 University Blvd
Birmingham, Alabama 35294-0022

@@

@@

__Objectives:__ Determine if the majority of US adults in subgroups defined by diabetes, chronic kidney disease (CKD), and age ≥65 years have high ASCVD risk (i.e., 10-year predicted ASCVD risk ≥10% or clinical CVD), and estimate the age-adjusted probability of having high ASCVD risk.

__Methods:__ Adults aged 40-79 years from the National Health and Nutrition Examination Survey `r min(exams)`-`r max(exams)+1` were included (n=`r table_value(nrow(current_analysis$data))`). We predicted 10-year ASCVD risk using the Pooled Cohort risk equations and clinical CVD was self-reported. Analyses were conducted overall and among those with stage 1 hypertension, defined by systolic BP of 130-139 mm Hg or diastolic BP of 80-89 mm Hg.

@@

@@

Distribution of US adults by blood pressure and antihypertensive medication use categories. In pie charts, the proportion of adults with stage 1 hypertension at high atherosclerotic cardiovascular disease risk in subgroups defined by diabetes, chronic kidney disease and age ≥ 65 years.
`r fp_par(line_spacing = 1)`

@@

knitr::include_graphics("../fig/central_illustration.png")

@@

High atherosclerotic cardiovascular disease risk was defined by a 10-year predicted risk for atherosclerotic cardiovascular disease ≥ 10% or clinical cardiovascular disease. Data are from the National Health and Nutrition Examination Survey, `r min(exams)`-`r max(exams)+1`.
`r fp_par(line_spacing = 1)`

@@

__Translational Outlook:__ Predicted 10-year ASCVD risk should be computed for all adults with stage 1 hypertension and without clinical CVD, as many are not at high risk for ASCVD.

@@

In November 2017, the `r ACC_AHA()` published a guideline for the prevention, detection, evaluation, and management of high `r BP()` in adults  [@whelton20182017]. This guideline recommends using both `r BP()` levels and predicted risk for `r CVD()` to guide the initiation of antihypertensive medication in addition to nonpharmacological therapy. Initiation of antihypertensive medication is recommended for all adults with a confirmed average `r SBP()` ≥ 140 mm Hg or `r DBP()` ≥ 90 mm Hg, stage 2 hypertension in the guideline. Initiation of antihypertensive medication is also recommended for adults with clinical `r CVD()` or a 10-year risk of `r ASCVD()` events ≥ 10% who have a confirmed average `r SBP()` between 130 and 139 mm Hg and/or `r DBP()` between 80 and 89 mm Hg, stage 1 hypertension in the guideline [@goff20142013]. Additionally, as a matter of practical convenience, adults with stage 1 hypertension and diabetes mellitus, `r CKD()`, or ≥ 65 years of age are considered to have high `r CVD()` risk by the guideline and initiation of antihypertensive medication is recommended. This recommendation, written in Sections 9.3, 9.6, and 10.3 of the guideline, was based on the assumption that the vast majority of adults with diabetes, `r CKD()`, or ≥ 65 years of age are likely to have a 10-year predicted `r ASCVD()` risk ≥ 10% and the knowledge that surveys suggest most clinicians, including cardiologists, do not assess `r CVD()` risk using recommended risk prediction tools [@whelton20182017; @shillinglaw2012use].

@@

The current analysis was restricted to adults aged 40 to 79 years of age who completed the `r NHANES()` interview and examination (n = `r n_age_40_to_79`). Participants < 40 or > 79 years of age were not included because use of the Pooled Cohort risk equations is not recommended in these age ranges [@goff20142013]. Participants who did not have three `r SBP()` and `r DBP()` measurements (n = `r n_bp_too_low`) and those who were missing information on age, race, sex, total and high-density lipoprotein cholesterol, smoking status, diabetes, or `r CKD()` status (n = `r n_missing_pcr`) were excluded. After these exclusions, a total of `r n_final` participants were included in the analysis (`r refer_to('fig_include_exclude')`).

@@

Data were collected during an in-home interview and a study visit completed at a mobile examination center. Standardized questionnaires were used to assess participants’ age, sex, race/ethnicity, smoking habits, medical history, use of antihypertensive medication, oral glucose lowering medication, and insulin. The medical history assessment included questions about whether the participant had been told by a doctor or other health professional that they had a heart attack, coronary heart disease, stroke, or heart failure. Clinical `r CVD()` was defined as answering yes to at least one of these questions.

@@

Blood and urine samples were collected during the medical examination. Serum creatinine, serum glucose and `r HB_A1C()` were measured using standard methods. Diabetes was defined by fasting plasma glucose ≥ `r gluc_cutpoint_fasted` mg/dL (≥ `r gluc_cutpoint_fed` mg/dL for those who were not fasting), `r HB_A1C()` ≥ `r hba1c_cutpoint`%, or self-reported use of insulin or oral glucose lowering medication. `r capitalize(E_GFR())` was calculated using the `r CKD()` Epidemiology Collaboration equation [@levey2009new; @cite_nephro]. Urinary albumin and creatinine levels were measured and used to calculate the `r ACR()`. `r CKD()` was defined by an `r E_GFR()` < `r egfr_cutpoint` ml/min/1.73m^2^ or an `r ACR()` ≥ `r acr_cutpoint` mg/g. Ten-year predicted `r ASCVD()` risk was calculated using the Pooled Cohort risk equations for participants without clinical `r CVD()` [@goff20142013; @lloyd2019use]. High `r ASCVD()` risk was defined as having a 10-year predicted `r ASCVD()` risk ≥ 10% or clinical `r CVD()`.

@@

Among US adults aged 40 to 79 years, the estimated percentage (95% `r CI()`) with high `r ASCVD()` risk was `r inline$prevHighRiskOverall` (`r refer_to('tab_risk_overall')`; bottom panel). Among sub-groups with diabetes, `r CKD()`, age ≥ 65 years, or any of these conditions, `r inline$prevHighRiskDiabetes`, `r inline$prevHighRiskCkd`, `r inline$prevHighRiskAge65`, and `r inline$prevHighRiskAny` had high `r ASCVD()` risk, respectively. Among those with stage 1 hypertension, the estimated percentage (95% `r CI()`) with high `r ASCVD()` risk was `r inline$prevHighRiskS1hOverall` in the overall population and `r inline$prevHighRiskS1hDiabetes`, `r inline$prevHighRiskS1hCkd`, `r inline$prevHighRiskS1hAge65`, and `r inline$prevHighRiskS1hAny` for those with diabetes, `r CKD()`, age ≥ 65 years, or any of these conditions, respectively.

@@

In the current study, the majority of US adults aged 40 to 79 years with diabetes, `r CKD()`, age ≥ 65 years, and any of these conditions had high `r ASCVD()` risk, defined by a 10-year predicted `r ASCVD()` risk ≥ 10% or clinical CVD. Also, the majority of US adults with stage 1 hypertension who were ≥ 65 years of age had high `r ASCVD()` risk. However, a substantial proportion of US adults with stage 1 hypertension and diabetes or `r CKD()` did not have high `r ASCVD()` risk. Among US adults with `r CKD()` and without high `r ASCVD()` risk, approximately one-third had a 10-year predicted `r ASCVD()` risk < 2.5%. In contrast, over 80% of US adults aged ≥ 65 years without high `r ASCVD()` risk had a 10-year predicted `r ASCVD()` risk between 5% and 10%. The probability of having high `r ASCVD()` risk was age-dependent with over 50% of US adults with diabetes and `r CKD()` having high `r ASCVD()` risk at ages above `r inline$ageHighRiskOvrlDiabetes` and `r inline$ageHighRiskOvrlCkd` years, respectively.

@@

Previous randomized trials and meta-analyses have investigated whether lower `r BP()` goals reduce incident `r CVD()` risk in patients with diabetes, `r CKD()`, or older age. A randomized trial of patients with diabetes and baseline `r SBP()` < 140 mm Hg and `r DBP()` < 90 mm Hg found lower risk of stroke among participants with more intense `r BP()` treatment goals [@schrier2002effects]. The Systolic Blood Pressure Intervention Trial compared an `r SBP()` treatment target of < 120 mm Hg to < 140 mm Hg among older adults (mean age of 68 years) without diabetes. Among participants with `r CKD()`, randomization to the lower `r SBP()` target reduced rates of major `r CVD()` events and all-cause death without evidence of effect modifications by `r CKD()` or deleterious effect on `r E_GFR()` or end stage renal disease [@cheung2017effects]. In the overall population, those randomized to the lower `r SBP()` target experienced lower rates of incident fatal and nonfatal major `r CVD()` events as well as all-cause mortality [@sprint2015randomized]. A systematic review and meta-analysis of 123 studies with 613,815 participants found strong support for lowering `r SBP()` to < 130 mm Hg and providing antihypertensive medication to adults with diabetes, CKD, and various other comorbidities [@ettehad2016blood]. Benefits associated with a `r SBP()` treatment goal of < 130 mm Hg have also been identified in several other meta-analyses [@bundy2017systolic; @reboussin2018systematic]. Collectively, evidence from previous studies suggests that adults with hypertension and diabetes, `r CKD()` or ≥ 65 years of age may experience lower rates of `r CVD()` events by initiating antihypertensive medication with an `r SBP()` treatment goal of < 130 mm Hg versus < 140 mm Hg.

@@

It has been suggested that most adults with diabetes, CKD, or ≥ 65 years of age have a 10-year predicted `r ASCVD()` risk ≥ 10% [@whelton20182017]. Previous research has shown that diabetes, `r CKD()`, and older age are each associated with an increased risk for `r ASCVD()` events [@coresh2004evidence; @chronic2010association; @grundy1999diabetes; @lakatta2002age]. Although the current study suggests that a high proportion of US adults with stage 1 hypertension and diabetes or `r CKD()` do not have a 10-year predicted `r ASCVD()` risk ≥ 10%, diabetes and `r CKD()` are associated with a high lifetime `r CVD()` risk [@lloyd2006prediction; @hippisley2010derivation]. Age-specific estimates in the current study suggest that US adults with diabetes or `r CKD()` develop high `r ASCVD()` risk at a younger age than their counterparts without these conditions. Prior studies have also found that cumulative exposure to high `r BP()` is associated with increased `r CVD()` risk [@allen2014blood]. Therefore, for younger adults with diabetes or `r CKD()`, early initiation of antihypertensive medication may be an important step towards lowering lifetime `r CVD()` risk. Estimating lifetime CVD risk may be useful when discussing the initiation of antihypertensive medication among young and middle-aged adults with diabetes or CKD.
